---
title: "Everybody's got to cheat sometimes...even me!"
author: "Xtophe Bontemps"
date: "2017-09-25"
output:
  html_notebook:
    code_folding: hide
    fig_caption: yes
    theme: journal
---

*A post from [data.visualisation.free.fr](http://data.visualisation.free.fr)*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, message =FALSE,results='hide'}
library(ggplot2)
library(ggthemes)
#library(doBy)
library(lubridate)
library(dplyr)
library(xtable)
library(readr)
library(reshape2)
library(ggthemes)

```

```{r data-processing, results='hide', echo=FALSE}
###### --- General ---- 
Myroot <- "c:/Chris/Cours2016/MOOC/"
#Myroot <- "F:/ZChris/Cours2016/MOOC/"

##### ----- USER Defined Course info----
# Course info
coursetitle <- "pricing-strategy-revenue-management"
courserun <- 1
courseSlug <- paste("pricing-strategy-revenue-management-",courserun, sep="")
courseStartDate <- as.Date("2017-01-16")

## Importing data

questionResponse <- read_csv(paste(Myroot,"DataReport-",courserun,"/data/", courseSlug, "_question-response.csv", sep=""), col_types = list(
  quiz_question = col_character(),
  response = col_character(),
  correct = col_character()
))

## ----Quizz and Test
quizbylearner <- questionResponse %>%
   group_by(learner_id) %>%
  summarise (
    nb_quiz= n_distinct(quiz_question)
      )

 SuccessbylearnerQuizz <- questionResponse %>%
  group_by(learner_id, quiz_question) %>% 
   summarize(
     attempts = n(),
     correct = sum(correct=="true"),
     score = abs(attempts-4)
   )%>%
   ungroup()
   
QuizzAnalysis <- merge(SuccessbylearnerQuizz , quizbylearner, by.x="learner_id")
 
QuizzAnalysis <- QuizzAnalysis %>% 
  group_by(learner_id) %>%
  mutate( 
      totalscore = sum(score),
      averagescore = totalscore/nb_quiz
              )

##-----TEST ANALYSIS ----####

# The Steps for our TEST are 1.15, 2.12, 3.21 , 4.4 and 4.10 !!!
TestAnalysis <-  mutate (SuccessbylearnerQuizz, step=substr(SuccessbylearnerQuizz$quiz_question, 1,4))%>%
  filter(step == "1.15" | step == "2.12"| step == "3.21"| step == "4.4."| step == "4.10" )%>%
  group_by(learner_id)  %>%
  mutate( nb_test = n_distinct(step) ) %>%
  group_by(learner_id, step) %>%
  mutate( test_score = sum(score)
                        )

#main source for test analysis (score for each learner, each test)
Scorebylearners <-  TestAnalysis %>%
  select(learner_id, step, nb_test, test_score )
Scorebylearners <-unique(Scorebylearners)  

#Nb of learners attendig the tests
LearnersbyTest <-  mutate (questionResponse, step=substr(questionResponse$quiz_question, 1,4))%>%
  filter(step == "1.15" | step == "2.12"| step == "3.21"| step == "4.4."| step == "4.10" )%>%
  group_by(step ) %>%
  summarize(
            Nb = n_distinct(learner_id)
  ) %>%
  ungroup()

#We add information on nb of leraners for each step
Scorebylearners <- merge(Scorebylearners, LearnersbyTest, by="step")
```


### Why ``cheating'' is sometimes useful in data science

Recently, I've worked on data from a MOOC we have created with some colleagues. The dataset was quite impressive since more than 3000 learners joined the course, viewed or interacted with some ressources (called ``*steps*''),  posted comments and pass some tests. One of our goal was to create a data visualisation that alowed us to see the results of the learners' tests, and, if possible, to detect some pattern in learners' results over the 5 tests. The data set looks like that:


```{r}
sample_n(select(Scorebylearners, learner_id, step, test_score), 10)
```


Using that dataset, we wanted to answer some questions:   

> Are  there some visible patterns? Are learners with good results for one test still good at another?  

So my first reflex was a plot with all the learners'results over the 5 steps:  

```{r point}
Plot.Point <- ggplot(Scorebylearners, aes(x=step, y= test_score)) + 
  geom_point(color = "grey", alpha=0.80) +
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Learners' score for each test ", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()

Plot.Point
```


Of course, the results to these tests are integers and take only some fixed values from 0 to 12 and many observations are *overlapping*.  

> This is a begginers' mistake! 

Well, so my second reflex was to use classical statistical representation such as the good old box-and-whiskers plot (boxplots)!  . 
```{r box }
Plot.Box <-  ggplot(Scorebylearners, aes(x=step, y= test_score)) + 
  geom_boxplot(outlier.colour= "grey", color= "darkgrey", fill="grey") + 
  guides(colour=FALSE, fill=FALSE)+
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Distribution of learners' score for each test (Box plot)", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()

Plot.Box
```

That's better, and I can see that there is some noticeable difference in the test results. 
But I wanted to see the individuals performances inside the boxes. 


> For that I have no choice but **to cheat a little bit ...**

### Cheating a little bit by adding some random noise...

In order to avoid overlapping, there are 2 basic tricks:
* to use *transparency* (or brushing, or alpha-transparency)
* to *jitter* the data by adding some random component to either the horizontal or vertical component. 

Let us add **transparency** and **horizontal jitter** only. 

  
```{r jitter.H }
Plot.Jitter.H <- ggplot(Scorebylearners, aes(x=step, y= test_score)) + 
  geom_jitter(color = "grey", alpha=0.20, width=0.20, height = 0) +
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Learners' score for each test (horizontal jitter)", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()
Plot.Jitter.H
```

> The points we see now (thanks to *jiter*) are not the original ones. Is that *cheating*?  

Let us do **transparency** and **vertical jitter**. 

  
```{r jitter.V, echo=FALSE, , results="hide"}
Plot.Jitter.V <- ggplot(Scorebylearners, aes(x=step, y= test_score)) + 
  geom_jitter(color = "grey", alpha=0.10, width=0, height = 0.20) +
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Learners' score for each test (vertical jitter)", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()
Plot.Jitter.V

```


Let us add **transparency** with  **horizontal and vertical jitter**. 

  
```{r jitter,  results="hide"}
Plot.Jitter <- ggplot(Scorebylearners, aes(x=step, y= test_score)) + 
  geom_jitter(color = "grey", alpha=0.60, width=0.40) +
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Learners' score for each test (Horizontal + vertical jitter)", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()
Plot.Jitter

```

### How *cheating*  helps for ploting lines in  parallel plots 

Now, if we want to follow learners results over time (over tests), we  can une *parallel plots* and draw lines linking each result.  

```{r SpaghettiBrut, results="hide"}
#Spaghetti plot original 
Plot.spaghetti <-  ggplot(Scorebylearners, 
                          aes(x=step, y= test_score,
                              group=factor(learner_id))) +
  guides(colour=FALSE) + 
  scale_x_discrete(name="Test step number", limits=c("1.15", "2.12", "3.21" , "4.4.", "4.10")) +
  scale_y_discrete(name ="Score",  limits=c(0,3,6,9, 12)) +
  labs(title = "Learners' score for each test  (Parallel plot)", 
       subtitle = paste("N=", nrow(Scorebylearners), "learners - ", nrow(TestAnalysis),"observations"),
       caption = "Source: MOOC ``Manage your prices'', FutureLearn (2017)"
       ) +
  coord_cartesian(ylim = c(0,12)) +
  theme_tufte()
 
#Plot Spaghetti brut
Plot.spaghetti + 
  geom_line( color="grey", size=1) +
  theme_tufte()


```

Since the score range from 1 to 12 and are discrete. many lines overlap and it is quite impossible to see some "pattern" in learners score. Nothing emerges from this simultation. 

> Let us *cheat* a little bit 

```{r SpaghettiJitter,  results="hide"}
#Adding jitter on Ys, and alpha-brushing 
Plot.spaghetti + 
  geom_line(alpha=0.30, color="grey", size=1, 
            aes(y = jitter(test_score, 2), x = step , group=factor(learner_id))) + 
  theme_tufte()


```

> Now we see it !

The difference between the two graph is quite striking. By adding some vertical noise on the Y axis - that is modifying randomly the  score value so that it is not integer any more - and using some *brushing* , help revealing some unseen and unnoticed patterns.  

We can also add some colour to follow individuals over those 


```{r SpaghettiColour,  results="hide"}
#
Plot.spaghetti + 
  geom_line(alpha=0.10, color=rainbow(nrow(Scorebylearners)), size=1.5, 
            aes(y = jitter(test_score, 2), x = step , group=factor(learner_id))) + 
  theme_tufte()



```





***
*Done in Toulouse (France), by [Xtophe](mailto:Xtophe.Bontemps@free.fr). Usual citation policy and disclaimer apply. Comments on my [twitter account](https://twitter.com/Xtophe_Bontemps) are welcome* 


